On Optimality of Finite Element Integration

### 2016-01-22 TOMS_Paper/assembly.tex
Various fixes all around

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, consider the operation $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector. One of the resulting sub-expressions will be $(a x^0_i + b x^1_i) (c x^0_i + d x^1_i)$, in which $x^0_i$ and $x^1_i$ represent components of $x$ depending on loop $L_i$. To schedule sub-expressions of this kind, we have two options:
\begin{enumerate}[label=(\roman*)]
\item Using generalized code motion -- not useful in the example as $x$ only depends on $L_i$.
\item Searching for spatial sharing, expanding, and then factorizing the expression to promote temporal sharing -- the expression in the example would be recast as $x^0_i (a + c) + x^1_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
\end{enumerate}
The reader can verify that (ii) improves the operation count over (i) independently of the size of $L_i$. In general, however, the optimal choice depends on multiple factors: the loop(s) size, the increase in operation count due to expansion, and the gain in code motion. We regard as \textit{structured} any tensor operation producing expressions for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of alternatives (i) and (ii). Of notable importance, from the perspective of achieving quasi-optimality, is the fact that all tensor operations involving test and trial functions are structured\footnote{It is also remarkable that structured tensor operations often characterize the \textit{entire} variational form. For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}.}. This observation is fundamental for understanding the sharing elimination algorithm.

\subsubsection{Algorithm}
\label{sec:se-algo}
Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest. It makes use of the following notation and terminology:
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
\item \textit{strategy (i)} and \textit{strategy (ii)}: the two approaches described in Section~\ref{sec:se-rln} for handling structured tensor operations.
\end{itemize}

@@

\item Perform a depth-first visit of the expression tree and, for each yet unhandled or hoisted sub-expression, apply the most profitable between strategies (i) and (ii). \\
\textit{Note: this pass speculatively assumes that structured tensor operations are in place. If the assumption does not hold, the result will generally be sub-optimal since only a subset of code motion opportunities may be considered.}
\end{enumerate}
\end{Algo}

@@

\section{Selection and composition of transformations}
\label{sec:optimal-synthesis}
In this section, we build a transformation algorithm that, given a memory bound, produces optimal or quasi-optimal finite element integration loop nests.

@@

Throughout the paper we have reiterated the idea that Algorithm~\ref{algo:gamma} reaches optimality if some preconditions are satisfied by the input variational form. We here discuss such preconditions, in increasing order of complexity.
\begin{enumerate}
\item There is a single monomial and only a specific coefficient, the coordinates field. This is by far the simplest scenario, which requires no particular transformation at the level of the outer loops, so optimality naturally follows.
\item There is a single monomial, but multiple coefficients are present. Optimality is achieved if only structured tensor operations are present (see Section~\ref{sec:se-rln}), since there is no ambiguity in factorization, which is exploited during step (7) of Algorithm~\ref{algo:sharing-elimination}.
\item There are multiple monomials, but no coefficients other than the coordinates field are present, or there is no spatial sharing in coefficients across different monomials. This reduces to cases (1) and (2) above.
\item There are multiple monomials and there is sharing of coefficients across different monomials. We can still in some cases ....
\end{enumerate}

@@

As discussed in the next section, all code generation systems used for performance evaluation handle block-sparse tables.

@@

Experiments were run on a single core of an Intel I7-2600 (Sandy Bridge) CPU, running at 3.4GHz, 32KB L1 cache (private), 256KB L2 cache (private) and 8MB L3 cache (shared). The Intel Turbo Boost and Intel Speed Step technologies were disabled. The Intel \texttt{icc 15.2} compiler was used. The compilation flags used were \texttt{-O3, -xHost, -ip}.

We analyze the runtime performance of four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods. In particular, we study the mass matrix (``\texttt{Mass}'') and the bilinear forms arising in a Helmholtz equation (``\texttt{Helmholtz}''), in an elastic model (``\texttt{Elasticity}''), and in a hyperelastic model (``\texttt{Hyperelasticity}''). The complete specification of these forms is made publicly available\footnote{\url{https://github.com/firedrakeproject/firedrake-bench/blob/experiments/forms/firedrake_forms.py}}.

@@

The values that we report are the average of three runs with ``warm cache'' (no code generation time, no compilation time). They include the cost of local assembly as well as the cost of matrix insertion. However, the unstructured mesh used for the simulations (details below) was chosen small enough to fit the L3 cache of the CPU so as to minimize the ``noise'' due to operations outside of the element matrix evaluation.

For a fair comparison, small patches (publicly available) were written to run \textit{all} simulations through Firedrake. This means the costs of matrix insertion and mesh iteration are identical in all variants. Our patches make UFLACS and the FEniCS Form Compiler's optimization systems generate code suitable for Firedrake, which employs a data storage layout different than that of FEniCS (e.g., array of pointers instead of pointer to pointers).

In Section~\ref{sec:mem-const}, we discussed the importance of memory constraints. We then define $T_H$ as the maximum amount of space that temporaries due to code motion can take. We set $T_H = L2_{size}$, that is, the size of the processor L2 cache (the last level of private cache). We recall that exceeding this threshold prevents the application of pre-evaluation. In our experiments, this happened in some circumstances. In such cases, experiments were repeated with $T_H = L3_{size}$ to verify the hypotheses made in Section~\ref{sec:mem-const}. We later elaborate on this.

@@

The rest of the section is structured as follows: we provide insights about the main message of the experimentation; we comment on the impact of autovectorization; we explain in detail, individually for each form, the performance results obtained.

\paragraph{High level view}
The main observation is that our transformation strategy does not always guarantee minimum execution time. In particular, 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime. The most significant of such test cases is the elastic model with $[q=4, nf=0]$. There are two reasons for this. Firstly, low level optimization can have a significant impact on actual performance. For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization. Secondly, memory constraints are critical, particularly the transformation strategy adpopted when exceeding $T_H$. We will later thoroughly elaborate on all these aspects.

@@

\paragraph{Hyperelasticity}
In the experiments on the hyperelastic model, shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} exhibits the largest gains out of all problem instances considered in this paper. This is a positive aspect: it means that our transformation algorithm scales with form complexity. The fact that all code transformation systems (apart from \texttt{tens}) show quite significant speed-ups suggests several points. Firstly, the baseline is highly inefficient: with forms as complex as in this hyperelastic model, a trivial translation of integration routines into code should always be avoided since even one of the best general-purpose compilers available (Intel's on an Intel platform at maximum optimization level) is not capable of exploiting the structure inherent in the mathematical expressions generated. Secondly, the code motion strategy really makes a considerable impact. The sharing elimination performed by \texttt{cfO2} in each level of the loop nest ensures a critical reduction in operation count, which results is better execution times. In particular, at higher order, the main difference between \texttt{ufls} and \texttt{cfO2} is due to the application of this transformation to the multilinear loop nest. Clearly, the operation count increases with $q$, and so do the speed-ups.

\section{Conclusions}
\label{sec:conclusions}
With this research we have set the foundation of optimal finite element integration. We have developed theory and implemented an automated system capable of applying it. The automated system, COFFEE, is integrated in Firedrake, a real-world framework for writing finite element methods. We believe the results are extremely positive. An open problem is understanding how to optimally handle non-linear loop nests. A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the use of high order function spaces. Lastly, we recall our work is publicly available and is already in use in the latest version of the Firedrake framework.

>>>>

\subsubsection{Identification and exploitation of structure}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, consider the operation $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector. The tensor operation will reduce to $(a x^0_i + b x^1_i) (c x^0_i + d x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ depending on loop $L_i$. To minimize the operation count for sub-expressions of this kind, we have two options:
\begin{enumerate}[label=(\roman*)]
\item Using generalized code motion -- not useful in the example as $x$ only depends on $L_i$.
\item Searching for spatial sharing, expanding, and then factorizing the expression to promote temporal sharing -- the expression in the example would be recast as $x^0_i (a + c) + x^1_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
\end{enumerate}
The reader can verify that (ii) improves the operation count over (i) independently of the size of $L_i$. In general, however, the optimal choice depends on multiple factors: the loop(s) size, the increase in operation count due to expansion, and the gain in code motion. We regard as \textit{structured} any sub-expression resulting from tensor operations for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of alternatives (i) and (ii). Of notable importance, from the perspective of achieving quasi-optimality, is the fact that sub-expressions involving test and trial functions are structured\footnote{It is also remarkable that structured sub-expressions often characterize the \textit{entire} variational form. For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}.}. This observation is fundamental for understanding the sharing elimination algorithm.

\subsubsection{Algorithm}
\label{sec:se-algo}
Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest. It makes use of the following notation and terminology:
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
\item \textit{strategy (i)} and \textit{strategy (ii)}: the two approaches described in Section~\ref{sec:se-rln} for handling structured sub-expressions.
\end{itemize}

@@

\item Perform a depth-first visit of the expression tree and, for each yet unhandled or hoisted sub-expression, apply the most profitable between strategies (i) and (ii). \\
\textit{Note: this pass speculatively assumes all sub-expressions to be structured. If the assumption does not hold, the result will generally be sub-optimal since only a subset of factorizations and code motion opportunities may be considered.}
\end{enumerate}
\end{Algo}

@@

\section{Selection and composition of transformations}
\label{sec:optimal-synthesis}
In this section, we build a transformation algorithm that, given a memory bound, systematically produces quasi-optimal finite element integration loop nests.

@@

Throughout the paper we have reiterated the idea that Algorithm~\ref{algo:gamma} reaches optimality if stronger preconditions on the input variational form are satisfied. We here discuss such preconditions, in increasing order of complexity.
\begin{enumerate}
\item There is a single monomial and only a specific coefficient (e.g., the coordinates field). This is by far the simplest scenario, which requires no particular transformation at the level of the outer loops, so optimality naturally follows.
\item There is a single monomial, but multiple coefficients are present. Optimality is achieved if and only if all sub-expressions depending on coefficients are structured (see Section~\ref{sec:se-rln}). This would avoid ambiguity in factorization, which in turn would guarantee the output of step (7) in Algorithm~\ref{algo:sharing-elimination} be optimality.
\item There are multiple monomials, but either at most one coefficient (e.g., the coordinates field) or multiple coefficients not inducing sharing across different monomials are present. This reduces, respectively, to cases (1) and (2) above.
\item There are multiple monomials, and coefficients are shared across monomials. Optimality is reached if and only if the coefficient-dependent sub-expressions produced by Algorithm~\ref{algo:sharing-elimination} -- that is, the by-product of factorizing test/trial functions from distinct monomials -- preserve structure. 
\end{enumerate}

@@

%As discussed in the next section, all code generation systems used for performance evaluation handle block-sparse tables.

@@

Experiments were run on a single core of an Intel I7-2600 (Sandy Bridge) CPU, running at 3.4GHz, 32KB L1 cache (private), 256KB L2 cache (private) and 8MB L3 cache (shared). The Intel Turbo Boost and Intel Speed Step technologies were disabled. The Intel \texttt{icc 15.2} compiler was used. The compilation flags used were \texttt{-O3, -xHost}.

We analyze the execution time of four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods. In particular, we study the mass matrix (``\texttt{Mass}'') and the bilinear forms arising in a Helmholtz equation (``\texttt{Helmholtz}''), in an elastic model (``\texttt{Elasticity}''), and in a hyperelastic model (``\texttt{Hyperelasticity}''). The complete specification of these forms is made publicly available\footnote{\url{https://github.com/firedrakeproject/firedrake-bench/blob/experiments/forms/firedrake_forms.py}}.

@@

The values that we report are the average of three runs with ``warm cache'' (no code generation time, no compilation time). They include the cost of both local assembly and matrix insertion, with the latter minimized through the choice of a mesh (details below) small enough to fit the L3 cache of the CPU.

For a fair comparison, small patches (publicly available) were written for compatibility of \texttt{quad}, \texttt{tens}, and \texttt{ufls} with Firedrake. By executing all simulations in Firedrake, we guarantee that matrix insertion and mesh iteration have a fixed cost independent of the transformation system employed. The patches, in particular, adjust the data storage layout (e.g., by generating array of pointers instead of pointer to pointers and by replacing flattened arrays with bi-dimensional ones, as Firedrake expects).

For Constraint~\ref{const:TH}, discussed in Section~\ref{sec:mem-const}, we set $T_H = L2_{size}$; that is, the size of the processor L2 cache (the last level of private cache). When the threshold had an impact on the transformation process, the experiments were repeated with $T_H = L3_{size}$. The results are documented later, individually for each problem.

@@

The rest of the section is organized as follows: we first provide insights into the overall outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we explain, in detail and individually for each form, the performance results obtained.

\paragraph{High level view}
The main observation is that our transformation strategy does not always guarantee minimum execution time. In particular, 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime. The most significant of such test cases is the elastic model with $[q=4, nf=0]$. There are two reasons for this. Firstly, low level optimization can have a significant impact on actual performance. For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization. Secondly, memory constraints are critical, particularly the transformation strategy adopted when exceeding $T_H$. We will later thoroughly elaborate on all these aspects.

@@

\paragraph{Hyperelasticity}
In the experiments on the hyperelastic model, shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} exhibits the largest gains out of all problem instances considered in this paper. This is a positive aspect, since it indicates that our transformation algorithm scales well with form complexity. The fact that all code transformation systems (apart from \texttt{tens}) show quite significant speed-ups suggests two points. Firstly, the baseline is highly inefficient. With forms as complex as in the hyperelastic model, a trivial translation of integration routines into code should always be avoided as even the best general-purpose compiler available (Intel's on an Intel platform at maximum optimization level) cannot exploit the structure inherent in the expressions. Secondly, the strategy for removing spatial and temporal sharing has a tremendous impact. Sharing elimination as performed by \texttt{cfO2} ensures a critical reduction in operation count, which becomes particularly pronounced for higher values of $q$.

\section{Conclusions}
\label{sec:conclusions}
We have developed transformation algorithms and a theory of optimality for finite element integration. The implementation was carried out in COFFEE, a compiler fully integrated with the Firedrake framework, and therefore made available to an increasingly significant user base. Performance results show the effectiveness of our methodology. There are still several open problems. The main one is an elaborate classification of forms for which optimality can be reached. Secondly, we would like to extend our methodology to classes of loops arising in spectral methods, in which interaction with low level optimization becomes even more critical due to the larger working sets.
